---
title: "Batch BFB Detection from Raw Copy-Number Profiles"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch BFB Detection from Raw Copy-Number Profiles}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width  = 10,
  fig.height = 6,
  out.width  = "100%",
  dpi        = 150,
  warning    = FALSE,
  message    = FALSE
)
```

```{r libraries}
library(bridges)
library(dplyr)
set.seed(1)
```

## Why a second detection method?

`detect_bfb_branches()` (see `vignette("inference")`) asks "does this *branch* of my inferred phylogeny look BFB-like?" -- it needs a fitted tree and ancestral reconstruction first. Sometimes you just want to ask a simpler, tree-independent question about a single cell's raw profile: **could this copy-number pattern have been produced by a BFB schedule at all?** That's an exact combinatorial question (Kinsella & Bafna 2012; Zakov, Kinsella & Bafna 2013), answered by `bfbtools`, and `bridges` wraps it into batch-friendly, `cna_data`-aware functions: `bfb_detect_vectors()`, `bfb_detect_batch()`, and `split_chr_arms()`.

Neither method supersedes the other -- they're complementary evidence.

---

## 1  From copy number to a count-vector

The BFB algorithms in `bfbtools` work on plain integer count-vectors, not on tibbles or matrices. Given one cell's copy-number row (already ordered along the chromosome), the count-vector is obtained by run-length-encoding it, dropping runs of zero (deletions), and requiring at least two remaining segments to be informative at all -- a flat single-segment row carries no BFB signal to test.

```{r extract-example}
row <- c(1, 1, 1, 0, 0, 3, 3, 5, 5, 5)
rle_vec <- rle(row)
v <- rle_vec$values
v <- v[v != 0]
v
```

`bfb_detect_batch()` (Section 3) does this extraction for you across an entire `cna_data` tibble; this section is just to show what's actually being tested underneath.

---

## 2  `bfb_detect_vectors()` -- the core, sim-agnostic batch logic

Given a plain list of count-vectors, `bfb_detect_vectors()` runs `bfbtools::admits_bfb()` on each, and `bfbtools::nearest_bfb()` for the ones that don't admit BFB exactly -- exploiting exact-duplicate profiles (very common across related cells in a lineage) so each distinct vector is only ever computed once.

```{r detect-vectors}
vec_list <- list(
  c(6, 3, 5),                   # admits BFB exactly
  c(6, 3, 5),                   # duplicate -- reused, not recomputed
  c(14, 7, 18, 16, 9, 12),      # a real near-miss: does NOT admit BFB exactly
  c(1, 1),                      # trivial: flat baseline + a deletion gap
  NULL                          # uninformative
)

bfb_detect_vectors(vec_list)
```

A few things worth noting in that output:

- **`bfb_distance`** is `-log(weight)` under a Poisson noise model: `0` for an exact match, increasing as more/larger corrections are needed to reach the nearest BFB-admitting vector.
- **The trivial vector `c(1, 1)` comes back `NA`, not `TRUE`.** "Admits a BFB schedule" is technically true even for zero BFB cycles -- the untouched baseline chromosome is itself a (trivial) BFB schedule. Since every real BFB cycle is an inverted *duplication*, it necessarily pushes some segment's copy number above 1; a profile with no amplification anywhere (just a flat baseline interrupted by a deletion) is definitionally uninteresting for this question. `min_max_count` (default `2`) enforces that precondition -- lower it to `1` only if you want the raw, weaker "technically admits BFB" semantics back.
- `n_unique` reports how many *distinct* count-vectors were actually run through `bfbtools` -- 3 here, out of 5 input rows.

---

## 3  `bfb_detect_batch()` -- straight from `cna_data`

`bfb_detect_batch()` extracts count-vectors from a `cna_data` tibble (one per cell, per chromosome, per allele) and calls `bfb_detect_vectors()` on all of them at once, so duplicate profiles across cells *and* alleles *and* chromosomes are deduplicated in a single pass.

```{r batch-sim}
sim <- bridge_sim(
  chromosomes = c(1:2), bfb_allele = "1:A",
  max_cells = 60, lambda = 1.5,
  bfb_prob = 0.5, amp_rate = 0.2, del_rate = 0.1
)

result <- bfb_detect_batch(sim$cna_data, alleles = c("A", "B"))
head(result)
```

The output has one row per `(cell_id, chr, allele)`: `n_segments` and `max_count` (useful complexity signals for building your own filters/weights beyond the built-in `min_max_count`), `admits_bfb`, `bfb_distance`, `nearest_weight`, and `n_unique`.

```{r batch-summary}
result %>%
  filter(!is.na(admits_bfb)) %>%
  count(chr, allele, admits_bfb)
```

Chromosome 1, allele A -- where BFB was seeded -- should show a higher `admits_bfb` rate than chromosome 2 or allele B.

---

## 4  Chromosome arms: why direction matters

A count-vector is only meaningful if it's oriented telomere-to-centromere. Real chromosomes have *two* telomeres, one on each arm, with the centromere in between -- so a whole-chromosome row mixes two opposite orientations into one, which will generally not read as a valid (or correctly-scored) BFB pattern even when the underlying biology is a clean BFB event confined to one arm.

`split_chr_arms()` labels each row `"p"` or `"q"` based on position relative to the centromere. Ascending genomic coordinate is already telomere-to-centromere on the p-arm, but is centromere-to-telomere (backwards) on the q-arm -- only the q-arm needs reversing before extraction, and `bfb_detect_batch()` handles this automatically once you supply a `genome`.

Centromere positions come from the `centromeres` dataset bundled with `bridges` -- real hg19 (GRCh37) and hg38 (GRCh38) coordinates for chromosomes 1-22, X, Y (see `?centromeres`). There is no manual-override or midpoint-guessing fallback: `genome` is the only way to enable arm-awareness, so arm-splitting is always anchored to real genomic coordinates for a real genome build.

```{r arm-aware}
# The SAME true [6,3,5] BFB pattern placed on both arms of chr1, each laid
# out in that arm's own natural raw coordinate order, at real-scale
# coordinates either side of hg19 chr1's real centromere (~121.5-124.5 Mb).
arm_cna <- rbind(
  data.frame(cell_id = "cellZ", chr = "1", start = c(100, 110, 120), end = c(109, 119, 129),
             A = c(6, 3, 5)),        # p-arm: ascending coordinate = telomere-first already
  data.frame(cell_id = "cellZ", chr = "1",
             start = c(124600000, 124600010, 124600020),
             end   = c(124600009, 124600019, 124600029),
             A = c(5, 3, 6))         # q-arm: ascending coordinate = centromere-first
)

bfb_detect_batch(arm_cna, alleles = "A", genome = "hg19")
```

Both arms correctly recover the same `[6, 3, 5]` pattern and both correctly admit BFB, despite having opposite raw coordinate orientation on disk -- because `genome` was supplied, `bfb_detect_batch()` ran `split_chr_arms()` internally, looked up real hg19 centromere positions, and applied the right reversal to each arm independently. Compare this to what a whole-chromosome (non-arm-aware) reading of the *same* six rows would see: a single 6-segment run that doesn't reduce to either arm's true pattern.

`genome` is one of `"hg19"` or `"hg38"`, or is left `NULL` by default (no arm-splitting; the whole chromosome is treated as one unit, matching the behavior in Section 3).

**A caveat worth being explicit about:** if `genome` is left `NULL` and your data spans both arms of a chromosome, whole-chromosome detection isn't just "less precise" -- it can silently give the wrong answer, in either direction, for exactly the reason shown above. If you're working with real (or realistically-simulated) whole-chromosome data, specifying `genome` is the safer default, not an optional refinement.

---

## Next steps

- **The tree-based alternative**: `vignette("inference")`.
- **The underlying algorithms on their own**, with no simulation or `cna_data` involved at all: `vignette("bfbtools-algorithms", package = "bfbtools")`.
