---
title: "Getting Started with bridges"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bridges}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.align = "center",
  fig.width  = 12,
  fig.height = 8,
  out.width  = "100%",
  dpi        = 150,
  warning    = FALSE,
  message    = FALSE
)
```

## Overview

`bridges` is an R package for inferring phylogenetic trees from single-cell copy number alteration (CNA) data, with explicit support for Breakage-Fusion-Bridge (BFB) mutational processes common in cancer.

**BFB in brief.** BFB cycles begin when a chromosome loses its telomere protection.  During replication the two sister chromatids fuse at their broken ends, forming a dicentric chromosome that is pulled to both poles during mitosis, causing a new break.  Repeated cycles amplify the genomic region near the original break, producing a characteristic "fold-back" copy number signature at one allele.

`bridges` provides:

- A **Gillespie simulator** (`bridge_sim`) for generating synthetic CNA data with configurable BFB, amplification, deletion, and whole-genome duplication events.
- An **inference pipeline** (`fit`) that computes allele-specific distance matrices and builds a neighbor-joining phylogenetic tree, penalizing BFB-like non-contiguous CN changes.
- **Downstream tools** for BFB detection (`detect_bfb`), visualization (`plot_heatmap`, `plot_bfb_signature`, `plot_chr_all_heatmap`), and ABC parameter inference (`abc_inference`).

---

## Input data format

All main functions expect a data frame with one row per cell-chromosome-bin combination:

| Column    | Type      | Description                              |
|-----------|-----------|------------------------------------------|
| `cell_id` | character | Unique cell identifier                   |
| `chr`     | character | Chromosome name (e.g. `"8"`, `"X"`)      |
| `start`   | integer   | Bin start position (bp)                  |
| `end`     | integer   | Bin end position (bp)                    |
| `CN`      | integer   | Total copy number (`A + B`)              |
| `A`       | integer   | Copy number of allele A (haploid)        |
| `B`       | integer   | Copy number of allele B (haploid)        |

---

## Step 1 — Simulate copy number evolution

`bridge_sim` runs a continuous-time Gillespie simulation.  Key parameters:

- `chromosomes` — which chromosomes to model.
- `bfb_allele` — which chromosome:allele undergoes BFB (e.g. `"8:A"`).
- `max_cells` — stop when this many cells are alive (evolutionary depth).
- `lambda` — mean number of genomic events per cell division.
- `bfb_prob`, `amp_rate`, `del_rate` — relative probabilities of each event type.

```{r simulate}
library(bridges)
library(ggplot2)
library(dplyr)

set.seed(42)
sim <- bridge_sim(
  chromosomes = c(1:8),
  bfb_allele  = "8:A",
  max_cells   = 5000,
  lambda      = 2, 
  subsample = 500,
  bfb_prob    = 0.85,
  amp_rate    = 0.2, 
  del_rate    = 0.1, 
  first_round_of_bfb = TRUE,
  hotspot = list(chr = "8:A", pos = 100)
)

# sim$cna_data is a long-format tibble ready for fit() and plotting
head(sim$cna_data)
```

The return value also includes `sim$tree` (ground-truth phylogeny), `sim$elapsed_time` (simulation clock), and `sim$n_alive`.

---

## Step 2 — Fit the phylogeny

`fit` preprocesses the CNA data, computes per-chromosome Greedy (G) and BFB/contiguity (B) distance matrices, merges them, and builds a neighbor-joining tree.

```{r fit}
res <- fit(
  data    = sim$cna_data,
  alleles = c("A", "B")
)
```

The result contains `res$tree`, `res$D` (final distance matrix), and `res$reconstructions` (ancestral CN states used for BFB detection).

---

## Step 3 — Compare inferred vs true tree

Robinson-Foulds distance measures topological dissimilarity between two trees. A value of 0 means identical topology; 1 means maximally different.

```{r compare-trees}
library(phangorn)

rf <- phangorn::RF.dist(sim$tree, res$tree, normalize = TRUE)
cat("Normalized RF distance:", round(rf, 3), "\n")
```

---

## Step 4 — Visualize CN profiles with the phylogeny

`plot_heatmap` combines a ComplexHeatmap of allele-specific CN profiles with the phylogenetic tree.

```{r plot-true-tree, fig.height=7}
# Ground-truth tree
plot_heatmap(
  sim$cna_data,
  tree         = sim$tree,
  to_plot      = c("CN", "A", "B"),
  ladderize    = TRUE,
  branch_length = 1,
  use_raster   = FALSE
)
```

```{r plot-inferred-tree, fig.height=7}
# Inferred tree with BFB-aware branch coloring
plot_heatmap(
  sim$cna_data,
  tree                 = res$tree,
  to_plot              = c("CN", "A", "B"),
  ladderize            = TRUE,
  branch_length        = 1,
  use_raster           = FALSE
)
```

---

## Step 5 — Detect BFB signatures

`detect_bfb` applies a pseudo-cell binomial test across all chromosomes and alleles to quantify the fraction of phylogenetic branches with BFB-like non-contiguous CN changes.  P-values are Benjamini-Hochberg corrected.

```{r detect-bfb}
bfb_df <- bridges:::detect_bfb(res, threshold = 0.005)
head(bfb_df)
```

```{r plot-bfb-bar, fig.height=4}
bfb_df %>%
  mutate(chr = factor(chr, levels = c(1:22, "X", "Y"))) %>%
  ggplot(aes(x = chr, y = mean, fill = allele)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0.1, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c(A = "#E64B35", B = "#4DBBD5")) +
  theme_bw() +
  lims(y = c(0, 1)) +
  labs(x = "Chromosome", y = "BFB-positive branch fraction", fill = "Allele",
       title = "BFB signal per chromosome")
```

Chromosome 8, allele A should show the strongest signal because that is where BFB events were seeded.  Use `plot_bfb_signature` for a segment-level view:

```{r plot-bfb-sig, fig.height=4}
plot_bfb_signature(res, chr_of_interest = "8", allele_of_interest = "A")
```

---

## Next steps

- **Simulator deep-dive** — BFB mechanics, selection, whole-genome duplication, clonal study tools, passage-series self-consistency, and ABC parameter inference: see `vignette("simulator")`.
- **Inference deep-dive** — distance metrics, jitter correction, BFB detection internals, advanced visualization: see `vignette("inference")`.
