---
title: "Phylogenetic Inference from CNA Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phylogenetic Inference from CNA Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width  = 11,
  fig.height = 7,
  out.width  = "100%",
  dpi        = 150,
  warning    = FALSE,
  message    = FALSE
)
```

```{r libraries}
library(bridges)
library(ggplot2)
library(dplyr)
set.seed(42)
```

---

## 1  Input data format

`fit()` and all plotting functions expect a data frame in long format with one row per **cell x genomic bin** combination. Required columns:

| Column    | Type      | Description                                   |
|-----------|-----------|--------------------------------------------------|
| `cell_id` | character | Unique cell identifier                        |
| `chr`     | character | Chromosome (e.g. `"1"`, `"X"`)               |
| `start`   | integer   | Bin start in base pairs                       |
| `end`     | integer   | Bin end in base pairs                         |
| `CN`      | integer   | Total copy number (`A + B`)                   |
| `A`       | integer   | Allele A copy number (haploid, diploid = 1)   |
| `B`       | integer   | Allele B copy number (haploid, diploid = 1)   |

**Loading real scWGS data.** Most single-cell WGS pipelines produce segment-level tables with similar columns. Rename columns to match the format above, convert chromosome names to bare numbers/letters (no `"chr"` prefix), and ensure bins are uniformly sized within each chromosome.

```{r setup-data}
# Generate synthetic data for this vignette
sim <- bridge_sim(
  chromosomes = c(1:2),
  bfb_allele  = "1:A",
  max_cells   = 256,
  lambda      = .75,
  bfb_prob    = 0.8,
  amp_rate    = 0.1,
  del_rate    = 0.1,
  subsample   = 256
)

# This is the data object we will use throughout
cna_data <- sim$cna_data
cat("Cells:", length(unique(cna_data$cell_id)), "\n")
cat("Chromosomes:", paste(unique(cna_data$chr), collapse = ", "), "\n")
```

---

## 2  The `fit()` pipeline

`fit()` runs a five-step pipeline:

1. **Validation** -- confirms distance functions exist and inputs are well-formed.
2. **Pre-processing** -- jitter correction (optional), matrix conversion, insertion of a synthetic diploid reference cell.
3. **Greedy (G) distances** -- per-chromosome, per-allele minimum-edit distance matrices.
4. **BFB/contiguity (B) distances** -- per-chromosome, per-allele distance matrices penalizing non-contiguous CN changes. Skipped when `b_dist_func = NULL`.
5. **Tree construction** -- `find_minimal_distances()` merges G and B matrices; `tree_func` (default `ape::nj`) builds the final tree.

```{r fit}
res <- fit(
  data    = cna_data,
  alleles = c("A", "B")
)
```

The result is a list:

```{r fit-names}
names(res)
```

- `res$tree` -- the inferred phylogenetic tree (class `phylo`, rooted at diploid).
- `res$D` -- final distance matrix across all cells.
- `res$greedy_Ds` / `res$avg_Ds` -- per-chromosome/allele matrices.
- `res$reconstructions` -- ancestral CN states per branch, used for tree-based BFB detection (Section 6).

---

## 3  Greedy (G) distance

The **Greedy distance** between two cells is the minimum number of integer copy-number steps needed to transform one CN profile into the other, computed independently for each chromosome and allele. It treats every bin independently, so a single event moving a large contiguous segment is counted the same as many independent changes.

Available G functions (inspect with `names(G_DISTS)`):

| Name            | Description                                         |
|-----------------|--------------------------------------------------------|
| `"greedy_fast"` | Per-bin absolute difference (default, fast)         |
| `"greedy_avg"`  | Averaged over alleles before summing                |

```r
# Greedy-only run (disable BFB modelling entirely)
res_g <- fit(data = cna_data, b_dist_func = NULL)
```

This is useful for ablation studies or when allele-specific data is unavailable.

---

## 4  BFB/contiguity (B) distance

The **BFB distance** adds a penalty for CN changes that are *non-contiguous* across genomic bins -- the hallmark of BFB amplification, where a fold-back creates a mirror-image pattern rather than a simple contiguous change. Combining G and B distances makes the final distance matrix sensitive to BFB-like evolutionary paths.

Available B functions (inspect with `names(B_DISTS)`):

| Name         | Description                                       |
|--------------|-------------------------------------------------------|
| `"bfb_fast"` | Contiguity penalty on the difference profile (default) |
| `"A_contig"` | Allele A contiguity                               |
| `"B_contig"` | Allele B contiguity                               |

`bfb_penalty` scales the contribution of the B distance relative to G. At the default of `0`, the two are combined symmetrically via `find_minimal_distances()`, which takes the element-wise minimum.

```{r b-dist}
# Increase BFB penalty to make the tree more sensitive to BFB events
res_bfb <- fit(data = cna_data, alleles = c("A", "B"), bfb_penalty = 1)
```

**When to set `b_dist_func = NULL`:** use greedy-only mode for total CN data (column `CN` only), for speed-ups in parameter sweeps, or as an ablation control when benchmarking BFB detection.

Note: `distances.R` also ships six older, pure-R implementations of these G/B distance functions (`find_greedy_distance()` and variants). They're kept for convenience/reference only -- `fit()`'s own defaults use the much faster C++ versions shown above -- and are slated for removal in a future version. If you're not intentionally selecting one of the old names, you're not using them.

---

## 5  Jitter correction

Real scWGS data often contains "jitter" -- small random fluctuations in CN values caused by segmentation noise. The `k_jitter_fix` parameter smooths profiles before distance computation:

- `k_jitter_fix = 0` -- no smoothing (default; appropriate for clean or simulated data).
- `k_jitter_fix > 0` -- a running median of width `2k + 1` is applied to each cell's CN profile before computing distances.

```{r jitter, eval=FALSE}
# For noisy real data, try k = 1 or 2
res_smooth <- fit(data = cna_data, alleles = c("A", "B"), k_jitter_fix = 1)
```

Start with `k_jitter_fix = 0` on simulated data and increase only if real data shows many small isolated CN changes that are likely artefactual.

---

## 6  BFB detection: two complementary methods

`bridges` provides two independent BFB detection mechanisms, answering different questions. This section covers the tree-based one, which is what `fit()`'s output feeds directly into; see `vignette("batch-detection")` for the other (which needs no tree or fit at all, and works on raw per-cell profiles using the exact combinatorial algorithms in `bfbtools`).

### The pseudo-cell approach

`detect_bfb_branches()` works on the reconstructed ancestral CN states stored in `res$reconstructions`. For each internal branch in the phylogeny, it constructs a **pseudo-cell** by comparing the CN profile of the ancestor with that of its child. It then computes:

- **G cost** -- greedy (contiguous) edit distance to explain the transition.
- **B cost** -- BFB/contiguity distance including the non-contiguous penalty.
- **delta** = G - B. A positive delta means the BFB model explains the transition *better* than simple edits: the CN change is non-contiguous, consistent with a fold-back event.

Branches where `delta > threshold` are counted as BFB-positive. A one-sided binomial test assesses whether the fraction of BFB-positive branches per chromosome-allele combination is significantly above chance. P-values are corrected with the Benjamini-Hochberg procedure.

```{r detect-bfb}
bfb_df <- detect_bfb_branches(res, threshold = 0.005)
bfb_df
```

Column interpretation:

- `mean` -- fraction of branches with BFB signal (between 0 and 1).
- `N_successes` / `N_trials` -- counts underlying the binomial test.
- `p.value` / `adj.pval` -- raw and BH-corrected p-values.
- `threshold` -- the delta threshold used.

Chromosome 1, allele A should have the highest `mean` because that is where BFB was seeded in our simulation.

### Tuning `threshold`

Lower threshold -> more sensitive (smaller delta counts as BFB), more false positives. Higher -> more specific. A value of `0.005` works well for simulated data; for noisy real data try `0.01`-`0.05` and inspect the distribution of delta values manually via `res$reconstructions`.

---

## 7  Visualization

### 7.1  `plot_heatmap()` -- combined CN + tree view

The primary output. Rows are cells ordered by the phylogenetic tree; columns are genomic bins.

```{r heatmap-basic, fig.height=6}
plot_heatmap(
  cna_data,
  tree                 = res$tree,
  chromosomes_to_plot  = c(1, 2),
  to_plot              = c("CN", "A", "B"),
  ladderize            = TRUE,
  branch_length        = 1,
  use_raster           = FALSE
)
```

**Reconstruction coloring** -- branches can be colored by BFB signal intensity on a chosen chromosome and allele:

```{r heatmap-recon, fig.height=6}
plot_heatmap(
  cna_data,
  tree                = res$tree,
  chromosomes_to_plot = c(1, 2),
  to_plot             = c("A", "B"),
  ladderize           = TRUE,
  branch_length       = 1,
  use_raster          = FALSE,
  reconstruction      = res$reconstructions,
  chr_for_coloring    = "1",
  allele_for_coloring = "A"
)
```

Darker / more saturated branches carry stronger BFB signal on chromosome 1, allele A.

**Other useful options:**

| Argument          | Effect                                                            |
|-------------------|------------------------------------------------------------------------|
| `reorder_tree`    | Optimize tip order to minimize crossed branches                   |
| `distance_matrix` | Distance matrix to use for tip reordering                        |
| `annotations`     | Data frame with `cell_id` + annotation columns -> row color bars   |
| `show_tree`       | `FALSE` to hide the tree (heatmap only)                           |

### 7.2  `plot_bfb_signature()` -- segment-level BFB map

```{r bfb-sig, fig.height=4}
plot_bfb_signature(res, chr_of_interest = "1", allele_of_interest = "A")
```

Each bar represents one genomic bin. Height = number of phylogenetic branches where that bin was involved in a BFB-like event. Bins near the original breakpoint accumulate the most events.

### 7.3  `plot_chr_all_heatmap()` -- per-chromosome detail

For a focused view of one chromosome without the phylogenetic tree:

```{r chr-heatmap, fig.height=5}
plot_chr_all_heatmap(
  cna_data,
  chr                   = "1",
  allele                = "A",
  order_heatmap         = TRUE,
  add_avg_CN_profile    = TRUE,
  add_gain_loss_profile = TRUE,
  use_raster            = FALSE
)
```

With `order_heatmap = TRUE`, rows are sorted by the number of bins matching the diploid baseline (fewest altered bins at the top). The annotation bars above show average CN (barplot) and gain/loss fractions (stacked proportions) per bin.

---

## Next steps

- **Tree-free BFB detection directly on raw profiles** (works without ever calling `fit()`): `vignette("batch-detection")`.
- **The underlying linear-time BFB algorithms on their own**: `vignette("bfbtools-algorithms", package = "bfbtools")`.
