---
title: "Simulating Copy Number Evolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Copy Number Evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width  = 10,
  fig.height = 6,
  out.width  = "100%",
  dpi        = 150,
  warning    = FALSE,
  message    = FALSE
)
```

```{r libraries}
library(bridges)
library(ggplot2)
library(dplyr)
set.seed(42)
```

---

## 1  The Gillespie model

`bridge_sim` implements a **continuous-time Gillespie algorithm** for cancer cell evolution.  At each step, the next event is chosen proportionally to total rates and the simulation clock is advanced by an exponential waiting time.

### Birth and death

| Parameter    | Meaning                                  | Default |
|--------------|------------------------------------------|---------|
| `birth_rate` | Rate at which a cell divides             | `0.1`   |
| `death_rate` | Rate at which a cell dies                | `0.001` |
| `max_cells`  | Stop when this many cells are alive      | `256`   |
| `max_time`   | Hard time limit                          | `300`   |

Cells with an amplified hotspot can have their birth rate boosted by `positive_selection_rate` (see Section 4).  The simulation stops as soon as `max_cells` **or** `max_time` is reached, whichever fires first.

### Genomic events per division

When a cell divides, each daughter independently draws a Poisson number of genomic events with mean `lambda`.  Each event is one of:

| Event          | Relative weight  | Description                          |
|----------------|------------------|--------------------------------------|
| BFB cycle      | `bfb_prob`       | Breakage-Fusion-Bridge on the target allele |
| Amplification  | `amp_rate`       | Focal copy gain on a random chromosome |
| Deletion       | `del_rate`       | Focal copy loss on a random chromosome |
| Normal dup     | `normal_dup_rate`| Whole-chromosome duplication (rarely used) |

The weights are automatically normalized to sum to 1.  For amplification and deletion events, the length of the affected segment is drawn from `Exp(1 / rate)` (in bins).

```{r sim-basic}
# A chromosome 7 simulation with moderate BFB and amplification
sim <- bridge_sim(
  chromosomes     = "1",
  bfb_allele      = "1:A",
  max_cells       = 200,
  lambda          = 2,
  bfb_prob        = 0.5,
  amp_rate        = 0.3,
  del_rate        = 0.2,
  birth_rate      = 1.0,
  death_rate      = 0.1
)

cat("Elapsed simulation time:", round(sim$elapsed_time, 2), "\n")
cat("Alive cells:", sim$n_alive, "\n")
cat("CNA tibble dimensions:", nrow(sim$cna_data), "rows x",
    length(unique(sim$cna_data$cell_id)), "cells\n")
```

The `elapsed_time` field records the simulation clock when the stopping criterion fired.  For `max_cells`-driven simulations with `birth_rate = 1.0`, elapsed time is approximately `log2(max_cells)` cell doublings.

---

## 2  BFB mechanics

### Specifying the BFB locus

The `bfb_allele` parameter identifies which chromosome:allele undergoes BFB cycles (e.g. `"1:A"` for allele A of chromosome 7).  The `hotspot` argument pinpoints the specific bin that is most likely to be amplified:

```{r}
x <- bridge_sim(
  chromosomes = "1",
  bfb_allele  = "1:A",
  hotspot     = list(chr = "1:A", pos = 60),  # bin 60 is the focal point
)
```

By default `pos = 100`, and the hotspot is used only for computing selection (Section 4) and for summary statistics.

### Breakpoint distributions

The breakpoint position within a BFB event is drawn from `breakpoint_support`:

| Value       | Behaviour                                                |
|-------------|----------------------------------------------------------|
| `"uniform"` | Uniform across the chromosome (default)                  |
| `"beta"`    | Beta distribution; requires `alpha` and `beta` params    |
| `"custom"`  | Sample from a user-supplied integer vector `custom_breakpoints` |

The `"custom"` mode is used by the ABC pipeline to match the observed breakpoint distribution in real data.

### `first_round_of_bfb`

When `TRUE` (default), the founding cell already undergoes one BFB event before the simulation starts.  This mimics a clone that was initiated by a single BFB event.  Set to `FALSE` to start from a fully diploid founder.

```{r bfb-effect, fig.height=4}
# Compare: with vs without first BFB round
s_with <- bridge_sim(chromosomes = "1", bfb_allele = "1:A",
                     max_cells = 100, lambda = 2, first_round_of_bfb = TRUE)
s_without <- bridge_sim(chromosomes = "1", bfb_allele = "1:A",
                        max_cells = 100, lambda = 2, first_round_of_bfb = FALSE)

cat("Max A copy number — with first BFB round:   ",
    max(s_with$cna_data$A, na.rm = TRUE), "\n")
cat("Max A copy number — without first BFB round:",
    max(s_without$cna_data$A, na.rm = TRUE), "\n")
```

---

## 3  Whole-genome duplication

Set `wgd_available > 0` to allow WGD events.  Each alive cell has a per-division probability `wgd_probability` of doubling its entire genome:

```{r wgd-example}
sim_wgd <- bridge_sim(
  chromosomes      = "1",
  bfb_allele       = "1:A",
  max_cells        = 100,
  lambda           = 1,
  wgd_available    = 1,      # allow up to 1 WGD per lineage
  wgd_probability  = 0.05
)

cat("Max total CN after potential WGD:",
    max(sim_wgd$cna_data$CN, na.rm = TRUE), "\n")
```

After a WGD event the typical diploid baseline doubles from 2 to 4 total copies.  Use `base_value = 2` in `summarise_clone()` when analyzing total CN data from WGD-exposed simulations.

---

## 4  Selective pressure

Cells that have gained at least one copy at the `hotspot` bin can be given a birth-rate advantage:

```{r selection-example}
# Neutral clone
sim_neutral <- bridge_sim(
  chromosomes = "1", bfb_allele = "1:A", max_cells = 150, lambda = 2,
  positive_selection_rate = 0
)

# Positively selected clone (amplified cells divide 3× faster)
sim_sel <- bridge_sim(
  chromosomes = "1", bfb_allele = "1:A", max_cells = 150, lambda = 2,
  positive_selection_rate = 3
)

cat("Neutral — median hotspot A CN:  ",
    median(sim_neutral$cna_data[sim_neutral$cna_data$chr == "1", ]$A), "\n")
cat("Selected — median hotspot A CN: ",
    median(sim_sel$cna_data[sim_sel$cna_data$chr == "1", ]$A), "\n")
```

`negative_selection_rate` penalizes cells *without* the hotspot gain, which has a similar net effect to positive selection but through a different mechanism (increased death rate for non-amplified cells).

---

## 5  Clonal study tools

The clonal simulation functions wrap `bridge_sim` for systematic comparisons across parameter combinations.

### 5.1  Single replicate

```{r simulate-clone}
p <- clonal_params(bfb_prob = 0.6, positive_selection_rate = 0)

res <- bridges:::simulate_clone(
  chromosome   = "1",
  allele       = "A",
  hotspot_pos  = 60,
  params       = p,
  max_cells    = 200,
  sample_cells = 50
)

cat("n_alive:", res$n_alive, "  n_cells (sampled):", res$n_cells, "\n")
summarise_clone(res$cna_matrix, hotspot_col = 60)
```

`simulate_clone` passes `sample_cells` directly to `bridge_sim(subsample = ...)`, so the CNA tibble is only ever built for the 50 kept cells — no double work.

### 5.2  Replicated simulations

`run_clonal_replicates` runs `N_replicates` independent clones and returns a tidy tibble of summary statistics:

```{r run-replicates}
reps <- run_clonal_replicates(
  chromosome   = "1",
  allele       = "A",
  hotspot_pos  = 60,
  params       = clonal_params(bfb_prob = 0.6),
  N_replicates = 10,
  max_cells    = 150,
  sample_cells = 50
)

dplyr::select(reps, hotspot_max_cn, max_cn_global, mean_breakpoints, n_alive) |>
  summary()
```

### 5.3  Multi-condition sweep

`compare_clonal_params` runs replicates for a named list of parameter sets and returns all results in one tibble.

```{r compare-conditions, fig.height=5}
conditions <- list(
  neutral       = clonal_params(positive_selection_rate = 0),
  bfb_strong    = clonal_params(bfb_prob = 0.8, positive_selection_rate = 0),
  bfb_selection = clonal_params(bfb_prob = 0.8, positive_selection_rate = 2)
)

sweep <- compare_clonal_params(
  param_list   = conditions,
  chromosome   = "1",
  allele       = "A",
  hotspot_pos  = 60,
  N_replicates = 10,
  max_cells    = 150,
  sample_cells = 50
)

plot_clonal_comparison(sweep,
  stats = c("hotspot_max_cn", "hotspot_cn_var", "max_cn_global", "mean_breakpoints"))
```

You can overlay observed statistics as reference lines by passing the output of `compute_observed_stats()` to the `observed_stats` argument.

---

## 6  Passage-series self-consistency

The passage-series framework tests whether the simulator correctly represents evolutionary accumulation: simulating more cells (longer elapsed time) should produce more complex CNA landscapes.

```{r passage-series, fig.height=5}
series <- simulate_passage_series(
  passage_cell_counts = c(early = 50, mid = 250, late = 5000),
  chromosome          = "1",
  allele              = "A",
  hotspot_pos         = 60,
  params              = clonal_params(),
  n_replicates        = 5,
  sample_cells        = 30
)

stats <- summarise_passage_series(series)
plot_passage_series(stats)
```

All complexity metrics (`hotspot_max_cn`, `max_cn_global`, `mean_breakpoints`, `hotspot_cn_var`) should increase monotonically from early → mid → late, confirming self-consistency.  Use `x = "passage_cells"` to show the raw max-cells value on the x-axis instead of expected doublings.

<!-- --- -->

<!-- ## 7  ABC parameter inference -->

<!-- `abc_inference` uses rejection Approximate Bayesian Computation to infer simulator parameters that match an observed CNA dataset.  It runs `n_simulations` independent forward simulations with parameters drawn from the prior, then accepts the closest `tolerance_quantile` fraction as the approximate posterior. -->

<!-- The six summary statistics compared are: per-bin gain rate, per-bin loss rate, mean CN profile, CN variance profile, maximum CN, and mean breakpoints per cell. -->

<!-- ```{r abc, eval=FALSE} -->
<!-- # Use the simulation as "observed" data (in practice, use real scWGS output) -->
<!-- obs <- bridge_sim( -->
<!--   chromosomes = "1", bfb_allele = "1:A", -->
<!--   max_cells   = 100,  lambda = 2, -->
<!--   bfb_prob    = 0.5,  amp_rate = 0.3, del_rate = 0.2 -->
<!-- ) -->

<!-- abc_res <- abc_inference( -->
<!--   cna_data           = obs$cna_data, -->
<!--   allele             = "A", -->
<!--   chromosome         = "1", -->
<!--   n_simulations      = 500,    # use >= 5000 in practice -->
<!--   tolerance_quantile = 0.05, -->
<!--   n_cores            = 1 -->
<!-- ) -->

<!-- abc_res$param_summary -->
<!-- ``` -->

<!-- ```{r abc-plot, eval=FALSE} -->
<!-- plot_abc_results(abc_res) -->
<!-- ``` -->

<!-- `plot_abc_results` displays a 7-panel figure: histograms of the approximate posterior for `amp_rate`, `del_rate`, `bfb_prob`, `death_fraction`, and `lambda`; the distribution of accepted distances; and a scatter of `amp_rate` vs `del_rate` coloured by `bfb_prob`. -->

<!-- **Practical guidance:** -->
<!-- - `n_simulations = 1000` with `tolerance_quantile = 0.01` gives 10 posterior samples — adequate for exploration. -->
<!-- - Use `n_simulations >= 5000` and `tolerance_quantile = 0.01–0.02` for publication-quality posteriors. -->
<!-- - Set `n_cores` to the number of available cores; each simulation is fully independent. -->
<!-- - The `pos` and `bin_length` arguments are automatically inferred from the observed data if not supplied. -->
