---
title: "Simulating Copy Number Evolution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Copy Number Evolution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.align = "center",
  fig.width  = 10,
  fig.height = 6,
  out.width  = "100%",
  dpi        = 150,
  warning    = FALSE,
  message    = FALSE
)
```

```{r libraries}
set.seed(42)
library(bridges)
```

This vignette covers the Gillespie simulation engine's own scope. If you're looking for serial-passage workflows or clonal-comparison summaries built on top of it, see `vignette("passages-and-clones")`.

---

## 1  The Gillespie model

`bridge_sim` implements a **continuous-time Gillespie algorithm** for cancer cell evolution. At each step, the next event is chosen proportionally to total rates and the simulation clock is advanced by an exponential waiting time.

### Birth and death

| Parameter    | Meaning                                  | Default |
|--------------|--------------------------------------------|---------|
| `birth_rate` | Rate at which a cell divides             | `0.1`   |
| `death_rate` | Rate at which a cell dies                | `0.001` |
| `max_cells`  | Stop when this many cells are alive      | `256`   |
| `max_time`   | Hard time limit                          | `300`   |

Cells with an amplified hotspot can have their birth rate boosted by `positive_selection_rate` (see Section 4). The simulation stops as soon as `max_cells` **or** `max_time` is reached, whichever fires first.

### Genomic events per division

When a cell divides, each daughter independently draws a Poisson number of genomic events with mean `lambda`. Each event is one of:

| Event          | Relative weight  | Description                          |
|----------------|------------------|----------------------------------------|
| BFB cycle      | `bfb_prob`       | Breakage-Fusion-Bridge on the target allele |
| Amplification  | `amp_rate`       | Focal copy gain on a random chromosome |
| Deletion       | `del_rate`       | Focal copy loss on a random chromosome |
| Normal dup     | `normal_dup_rate`| Whole-chromosome duplication (rarely used) |

The weights are automatically normalized to sum to 1. For amplification and deletion events, the length of the affected segment is drawn from `Exp(1 / rate)` (in bins).

```{r sim-basic}
# A chromosome 1 simulation with moderate BFB and amplification
sim <- bridge_sim(
  chromosomes     = "1",
  bfb_allele      = "1:A",
  max_cells       = 200,
  lambda          = 2,
  bfb_prob        = 0.5,
  amp_rate        = 0.3,
  del_rate        = 0.2,
  birth_rate      = 1.0,
  death_rate      = 0.1
)

cat("Elapsed simulation time:", round(sim$elapsed_time, 2), "\n")
cat("Alive cells:", sim$n_alive, "\n")
cat("CNA tibble dimensions:", nrow(sim$cna_data), "rows x",
    length(unique(sim$cna_data$cell_id)), "cells\n")
```

The `elapsed_time` field records the simulation clock when the stopping criterion fired. For `max_cells`-driven simulations with `birth_rate = 1.0`, elapsed time is approximately `log2(max_cells)` cell doublings.

---

## 2  BFB mechanics

### Specifying the BFB locus

The `bfb_allele` parameter identifies which chromosome:allele undergoes BFB cycles (e.g. `"1:A"` for allele A of chromosome 1). The `hotspot` argument pinpoints the specific bin that is most likely to be amplified:

```{r hotspot}
x <- bridge_sim(
  chromosomes = "1",
  bfb_allele  = "1:A",
  hotspot     = list(chr = "1:A", pos = 60)  # bin 60 is the focal point
)
```

By default `pos = 100`, and the hotspot is used only for computing selection (Section 4) and for summary statistics.

### Breakpoint distributions

The breakpoint position within a BFB event is drawn from `breakpoint_support`:

| Value       | Behaviour                                                |
|-------------|------------------------------------------------------------|
| `"uniform"` | Uniform across the chromosome (default)                  |
| `"beta"`    | Beta distribution; requires `alpha` and `beta` params    |
| `"custom"`  | Sample from a user-supplied integer vector `custom_breakpoints` |

`"custom"` mode only applies to the founding cell's initial BFB event (`first_round_of_bfb`, below) -- the main simulation loop (implemented in C++ for speed) supports `"uniform"` and `"beta"` only. If you need custom breakpoints, combine `breakpoint_support = "custom"` with `bfb_prob = 0` so no further BFB events are attempted after the founding one:

```{r custom-breakpoints}
x_custom <- bridge_sim(
  chromosomes = "1",
  bfb_allele  = "1:A",
  max_cells   = 50,
  bfb_prob    = 0,                      # no further BFB after the founding event
  first_round_of_bfb = TRUE,
  breakpoint_support = "custom",
  custom_breakpoints = c(50, 150, 250)
)
```

### `first_round_of_bfb`

When `TRUE` (default), the founding cell already undergoes one BFB event before the simulation starts. This mimics a clone that was initiated by a single BFB event. Set to `FALSE` to start from a fully diploid founder.

```{r bfb-effect, fig.height=4}
# Compare: with vs without first BFB round
s_with <- bridge_sim(chromosomes = "1", bfb_allele = "1:A",
                     max_cells = 100, lambda = 2, first_round_of_bfb = TRUE)
s_without <- bridge_sim(chromosomes = "1", bfb_allele = "1:A",
                        max_cells = 100, lambda = 2, first_round_of_bfb = FALSE)

cat("Max A copy number — with first BFB round:   ",
    max(s_with$cna_data$A, na.rm = TRUE), "\n")
cat("Max A copy number — without first BFB round:",
    max(s_without$cna_data$A, na.rm = TRUE), "\n")
```

---

## 3  Whole-genome duplication

Set `wgd_available > 0` to allow WGD events. Each alive cell has a per-division probability `wgd_probability` of doubling its entire genome:

```{r wgd-example}
sim_wgd <- bridge_sim(
  chromosomes      = "1",
  bfb_allele       = "1:A",
  max_cells        = 100,
  lambda           = 1,
  wgd_available    = 1,      # allow up to 1 WGD per lineage
  wgd_probability  = 0.05
)

cat("Max total CN after potential WGD:",
    max(sim_wgd$cna_data$CN, na.rm = TRUE), "\n")
```

After a WGD event the typical diploid baseline doubles from 2 to 4 total copies. If you use `bridges::summarise_clone()` downstream on total CN data from WGD-exposed simulations, pass `base_value = 2`.

---

## 4  Selective pressure

Cells that have gained at least one copy at the `hotspot` bin can be given a birth-rate advantage:

```{r selection-example}
# Neutral clone
sim_neutral <- bridge_sim(
  chromosomes = "1", bfb_allele = "1:A", max_cells = 150, lambda = 2,
  positive_selection_rate = 0
)

# Positively selected clone (amplified cells divide 3x faster)
sim_sel <- bridge_sim(
  chromosomes = "1", bfb_allele = "1:A", max_cells = 150, lambda = 2,
  positive_selection_rate = 3
)

cat("Neutral — median hotspot A CN:  ",
    median(sim_neutral$cna_data[sim_neutral$cna_data$chr == "1", ]$A), "\n")
cat("Selected — median hotspot A CN: ",
    median(sim_sel$cna_data[sim_sel$cna_data$chr == "1", ]$A), "\n")
```

`negative_selection_rate` penalizes cells *without* the hotspot gain, which has a similar net effect to positive selection but through a different mechanism (increased death rate for non-amplified cells). `selection_type` controls how hotspot copy number translates into that advantage: `"constant"` (default; any gain above baseline gives the same fixed advantage), `"linear"` (advantage scales with copy number, unbounded), or `"saturation"` (advantage saturates, controlled by `saturation_K`).

---

## Next steps

- **Serial passages and clonal summaries** (repeated bottlenecked growth, summary statistics): `vignette("passages-and-clones", package = "bridges")`.
- **Fitting a phylogeny from simulated or real data**: `vignette("inference", package = "bridges")`.
- **Detecting BFB signal**, either from a fitted tree or directly from raw profiles: `vignette("bridges", package = "bridges")` for the overview, or `vignette("batch-detection", package = "bridges")` for the tree-free method.
